---
title: "Introduction to R"
author: "Jessica Kosie"
date: "9/17/2018"
output:
  pdf_document:
    toc: yes
  html_document:
    toc: yes
    toc_float: yes
  word_document:
    toc: yes
---

# Goals 

By the end of this tutorial, you will know:

+ Basic `R` usage (using `R` as a calculator, creating variables, indexing vectors)
+ How to read in and examine data
+ How to get values out of the rows and columns in your data
+ What a pipe is and how to use pipes to chain together `tidyverse` verbs
+ How to create useful summaries of your data using tidyverse

The best way to do this tutorial is to walk through it slowly, executing each line and trying to understand what it does. You can execute a single line by placing your cursor somewhere in that line and hitting CMD+enter (again on a mac) and execute a whole chunk at a time by hitting CMD+option+C (on a mac). 

# 1. Basic R Use

R can simply be used as a calculator.

```{r}
# Basic arithmetic
2+3
2*3
10/2
4^2

# Follows order of operations (PEMDAS)
(2^3)+4*(5/3)
```

These values aren't stored anywhere though. 

To keep them in memory, we need to assign them to a variable. 

```{r}
# Create a variable called x, that is assigned the number 8. 
x <- 8 
x = 8

# What value did I assign to x?
x

# What about y?
y

# Need to assign something to y!
y <- 2

y


# And we can do things with these variables:
x + y
x * y

z <- x * y

# Can use any name you want! But...object names must start with a letter, and can only contain letters, numbers, "_" and "." So, there are different options.

this_is_snake_case <- 8
thisIsCamelCase <- 8
some.people.use.periods <- 8
And_someTimes.People_doSTRANGEtHiNgS <- 8

# Can also assign a range of values
x <- 1:5 #x is now a vector of 1, 2, 3, 4, 5

# Note that x is no longer 8, it take on whatever the most recent assignment was

# We can also assign a vector of values this way
x <- c(2, 8, 1, 9)
```

Vectors are just 1-dimensional lists of numbers.

```{r}
# Let's get the numbers 1 thru 10 by ones 
1:10

# Sequence of numbers, 1 thru 10, by 2
seq(from = 1, to = 10, by = 2) #can say "from", "to", and "by=", but they're not necessary

```

> **Exercise 1a.** Create a variable called x that is assigned the number 8. Create a variable called y that is a sequence of numbers from 2 to 16, by 2. Add x and y. What happens? 

```{r}
x <- 8
y <- seq(from = 2, to = 16, by = 2)
x+y

```

> **Exercise 1b.** Create a variable called z that is a vector containing the numbers 5 and 10. Create a variable called q that is a sequence of numbers from 5 to 25, by 5. Multiply x and y. What happens? 

```{r}
x <- c(5, 10)
y <- seq(from = 5, to = 25, by = 5)
x*y

# Note: TODO add warnings vs. errors in file w/ answers 

```
## 2. Functions

`seq` that we used above is a **function**. Everything you typically want to do in statistical programming uses functions. Functions can take different **arguments**.

For example, `sqrt` is a function that can take one numeric argument. We are going to **apply** the function `sqrt` to the value 16.
```{r}
sqrt(16)

# I could have assigned 16 to a variable first:
x <- 16

# And then taken the square root of that variable:
sqrt(x)

```
The function `sqrt` can also take a vector as the argument.
```{r}
# Create a vector:
x <- 1:10

# Get the square root of every number in that vector:
sqrt(x)

```
Another good example is the function `mean`. `mean` takes one **argument**, a numeric vector. We are going to create a vector called `z` and **apply** this function to that vector.

```{r}
z <- 0:20
mean(z)

# Now, let's get the mean of this vector:
q <- c(2, 8, 6, NA, 4, 8)

q

mean(q)

# Is that the answer you'd expect? let's get some information about the function `mean`
?mean

# We need to add an additional argument to tell the function that we want to ignore NAs (the default for this argument is FALSE, that's why NAs weren't ignored above)
mean(q, na.rm = TRUE)
```

> **Exercise 2a.** R has a function called `rnorm` that will allow you to get a random sample of numbers drawn from a normal distribtuion. Get a sample of 5 numbers with a mean of 0 and a standard deviation of 0.5.  

```{r}
# Hint:
?rnorm

x <- rnorm(5, sd = 0.5)
```

There are multiple ways to create arrays of number.

```{r}
# As before, we can get a sequence of numbers, 1 thru 10, by 2
s <- seq(1,10,2) #can say "by=" before the 2, but it's not necessary

# Sequence of 11 equally spaced numbers between 0 and 1
s2 <- seq(0,1,length.out=11) 

# Side note: what if I don't want these things in my environment any more?
rm(x)
rm(q)
rm(s)

# Remove everything
rm(list=ls())

# Can bind things together

x <- 1:10
y <- 11:20
z <- c(x,y) #combines them into one vector
z1 <- cbind(x,y) #combines them into two columns in a matrix
z2 <- rbind(x,y) #combines them into two rows in a matrix

```

Speaking of matrices: Creating and indexing matrices

```{r}
x <- matrix(c(11,12,13,21,22,23), byrow=TRUE,nrow=2) #put into a matrix by row
x1 <- matrix(c(11,12,13,21,22,23), byrow=FALSE,nrow=2) #put into a matrix by column

# Indexing matrices (getting the value that's in a particular row/column)

# x[r,c] would give you the element in row r, column c of the matrix

x[2,3] #gives you the element in row 2 column 3 of matrix x (defined above)
x[] #all rows, all columns - could have just typed x
x[1, ] #first row, all columns
x[ ,3] #all rows, third column
x[1,3] #first row, third column

y <- x[1,2] #assign the value in the 1st row 2nd column to a new variable
```

> **Exercise 2b.** Create a matrix with the following values:
row 1: 10, 26, 32
row 2: 4, 8, 16
row 3: 9, 12, 10

Save it to the variable my_matrix.

```{r}

my_matrix <- matrix(c(10, 26, 32, 4, 8, 16, 9, 12, 10), byrow = TRUE, nrow = 3)

```

> **Exercise 2c.** Use indexing (as described above) to pull out the value in row 1, column 3. What is this value? 

```{r}

my_matrix[1,3]

```

> **Exercise 2d.** Next, using indexing to pull out the value 16. (Hint: which row and column is it in?) Save it to the variable z.

```{r}

z <- my_matrix[2,3]

```

> **Exercise 2e.** Finally, save column 2 to a variable called col_2 and add this to z.

```{r}

col_2 <- my_matrix[,2]

col_2 + z

```

## 3. Generating random numbers.

```{r}

# Previously we used `rnorm`, `runif` is similar, but it draws from a uniform distribution.

runif(1) #generates 1 random number from a uniform distribution from 0 to 1 (as defaults)
runif(2, min=3, max=5) #generates 2 random numbers from a uniform distribution from 3-5

#can hit the up arrow and run this again and again - draw lots of samples of 2
#can hit the up arrow and change things:
#up arrow gives you this runif(2, min=3, max=5) if you haven't yet run the line, you can edit any of the values.

# Sample gives whole numbers
sample(1:10, 3) #without replacement
sample(1:10, 3, replace=TRUE) #keep running this you might get a couple #s that are the same
sample(1:10, 30, replace=TRUE) #more likely to get numbers that are the same since it's a bigger sample

# Note that we'll get different numbers, unless we all set the same seed before we run the function
set.seed(5678)
sample(1:10, 3)

# Can create a matrix of random numbers
x <- matrix(runif(100), ncol=2) #100 row, 2 column vector of random numbers from a uniform dist.

x <- rnorm(5) #5 numbers from a normal dist. w/ mean=0 SD=1

# Can set parameters of the normal distribution
x <- rnorm(1, mean=100, sd=15) #draw one number from normal dist. w/ mean=100, SD=15 (e.g., 1 IQ score)

# We can also round that number
round(x, 2)

```

> **Exercise 3a.** Set the seed to 8642 and create a matrix (with 10 rows and 20 columns) of 200 numbers randomly drawn from a normal distribution. Use the default mean and sd. Enter the numbers by row. Save this matrix to a variable (you can call it anything you'd like).  

```{r}

set.seed(8642)
anything_you_like <- matrix(rnorm(200), ncol=20, byrow = TRUE) 

```

> **Exercise 3b.** Use indexing to get the value in row 2, column 17. Round this value to three decimal places.  

```{r}

x <- anything_you_like[2,17]
round(x, 3)

# or

round(anything_you_like[2,17], 3)

```

# 4. Plotting in "Base R"

```{r}
# Get some data and plot it
plot(sin(seq(-pi,pi,.01))) #plot the sin of every value in the sequence between -pi and pi

# Plot 1 to 10
plot(seq(1,10,1))

# Plot that as a line
plot(seq(1,10,1), type='l')

# Plot that as a blue line
plot(seq(1,10,1), type='l', col = "blue")

# We can plot 2 graphs on the same plot

# Get some data 
x <- seq(-pi,pi,.01)

# Create one plot
plot(sin(x))

# Keep the plot open to overlay something else
par(new=TRUE)

# Plot another graph in a different color so we can easily see the two different graphs
plot(cos(x), col="green")

# Here's another way to do it
lines(tan(x),col="blue") #puts things in same plot, but doesn't overlay

# Get a different x
x <- rnorm(100, mean=100, sd=15) #100 values from normal dist. w/ mean=100, SD=15
#can check mean and sd of our sample
mean(x)
sd(x)
#pretty close to the parameters we set because the sample size is large 
#probably wouldn't be as close if sample size was 5

# Get a histogram of x
hist(x)

# Can do fancier stuff
hist(x, breaks=20) # histogram of x with 20 bins
hist(x, breaks=20, xlab="IQs", main="Histogram of IQs") #title and axes

# Line graphs
x <- seq(-3,3,.01)
y <- seq(1,7,.01)

# Can give values for the x and y axis
plot(x,y)

# Can do fancier stuff here as well
plot(x, y, type = 'l', col = "purple", xlab = "X Axis", ylab = "Y Axis", main = "Graph of X and Y")

# Now let's look at bar plots - get some categorical data
x <- c(10,8,7,7,5)
barplot(x,main="Some RTs", xlab="Group")

# Label the bars
barplot(x, main="Some RTs", xlab="Group", names.arg=c("One", "Two", "Three", "Four", "Five"))

# Add some color
barplot(x,main="Some RTs", xlab="Group", names.arg=c("One", "Two", "Three", "Four", "Five"), col=rainbow(5))

# Which arguments did we use?
?barplot

# Add a legend (probably don't really need this here since bars are labeled...)
legend("topright", c("One", "Two", "Three", "Four", "Five"), cex=0.6, bty="n", fill=rainbow(5))

# What do these arguments mean
?legend

```

# For simple plotting in Base R, I really like this website: https://www.harding.edu/fmccown/r/

(below are some examples borrowed from this website)

This site is nice, too, for understanding graphical parameters: https://www.statmethods.net/advgraphs/parameters.html

```{r}
# Compare cars and trucks
cars <- c(1,3,6,4,9)
trucks <- c(2,5,4,5,12)

# Make a graph for cars. using points connected by a line, y axis ranges from 0 to 12.
plot(cars, type='o', col="blue", ylim=c(0,12), ann=FALSE) 

#ann=FALSE means that R shouldn't label the axes, asking it not to automatically label the y axis right now because it would just be labeled cars and we want to plot trucks, too.

# Graph trucks on the same plot w/ a red dashed line and square points
lines(trucks, type='o', pch=22, lty=2, col="red")

# Add a title & labels for axes
title(main="Autos",col.main="red",font.main=4, xlab="Count", ylab="Total")

```

> **Exercise 4a.** Let's create a graph comparing number of yoga classes taken in a week and stress level. 

Create a variable called `yoga` containing the numbers 0, 1, 2, 3, 4, 5, 6, 7. 

Create a second variable called `stress` containing the numbers 10, 8, 7, 4, 3, 2, 3, 8.  

Create a plot with `yoga` on the x axis and `stress` on the y axis.   

Make sure this is a line graph and make your line blue and dashed. 

Label the x axis `Number of Yoga Classes`.

Label the y axis `Stress Level`. Make sure the y axis ranges from 0 to 10.

Title the graph `Relationship between Yoga and Stress`.

What is the relationship between yoga and stress?
```{r}

yoga <- c(0, 1, 2, 3, 4, 5, 6, 7)

stress <- c(10, 8, 7, 4, 3, 2, 3, 8)

plot(yoga, stress, type = 'l', lty = 2, col = 'blue', ylim = c(0,10), xlab = 'Number of Yoga Classes', ylab = 'Stress Level', main = 'Relationship between Yoga and Stress')

```

## 5. Reading data into R

First, you'll need to tell R where to look for the data. To do this, you will set your working directory. 

For this tutorial, your working directory should be wherever you downloaded the materials. (They should be on your desktop.)

```{r}

# What is my current working directory?
getwd()

# How can I change my working directory?
setwd("/Users/jessicakosie/Desktop/uoregon_r_bootcamp/")
getwd()

# What is in my working directory?
dir()

```

You can also do this using RStudio, via the graphical interface on the top:

Session > Set Working Directory > Choose Directory

You can choose the folder you want to work in. The code for setting the working directory will populate in the console. You can then copy/paste this into your code if you'd like.

```{r}
# Now, let's read in the pragmatic_scales_data CSV file and save it as an object called ps_data
ps_data <- read.csv("pragmatic_scales_data.csv", header = TRUE)

```

> ProTip: You can also read in data files generated by other programs like SPSS. There are often packages to help with this.

Here's how you could read in this file if it were an SPSS file. (The code below won't work because the file is a .csv)
```{r}
# Uncomment the line below to install the `foreign` package
# install.packages("foreign")
#library(foreign)
#ps_data_spss <- read.spss('pragmatic_scales_data.sav', to.data.frame = TRUE)

```

# Examining the data file

You should now see the dataset in your environment.

We can simply look at the data frame. We can also get a summary of the data. (these are all **functions** too!)

```{r}
# Look at the first few rows of the data.
head(ps_data)

# Look at the final few rows of the data.
tail(ps_data)

# You can also specify the number of rows you'd like to see
head(ps_data, n = 10)
tail(ps_data, n = 2)

# Get a summary of the data.
summary(ps_data)

# You can also `View` data, which shows an interactive "spreadsheet" view. You can also get this by clicking on the dataset in the environment.
View(ps_data)

# Note the capital `V`
view(ps_data) #this doesn't work!

# We can also look at the structure of the data.
str(ps_data)

# I can get the column names of the data.
colnames(ps_data)

```

> **Exercise 5a.** Move the file called `another_data_set.csv` from the `uoregon_r_bootcamp` folder on your desktop to a different folder on your computer. Re-set your working directory to that folder. Use `getwd()` to make sure you have the right directory. Read in this new dataset and name it anything you'd like. Use head/tail/summary/etc to check out this data set.

Once you've read in that data file, reset your working directory to the `uoregon_r_bootcamp` folder on your desktop.
```{r}

# I moved the file to `Documents`

setwd("/Users/jessicakosie/Documents")
getwd()
more_data <- read.csv("another_data_set.csv", header = T)

setwd("/Users/jessicakosie/Desktop/uoregon_r_bootcamp/")
getwd()

```

## Playing around with SWIRL

```{r}

# I highly recommend trying out the swirl package!

install.packages("swirl") # give it time to install, if you're using R Studio you'll see a little 
# stop sign in the top right corner of the console until it's finished installing.
library(swirl)
swirl()  # have fun! :)

```
